Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods

### 2022-12-24 inst/surfing_jsa.tex
issue #166

\documentclass{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{diagbox}
\usepackage{multirow}
\usepackage[pdftex]{hyperref}

@@

Figure~\ref{dotchartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about $0.175$ down to Morais at about $0.01$.  The {\tt
  hyper2} software~\citep{hankin2017_nomarkup} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  This test is performed by function {\tt
  equalp.test()} in the package which uses the method of support
\citep{edwards1992}.  The test rejects the null (in favour of the
evaluate) at $p=4\times 10^{-5}$.  There is thus strong evidence to
suggest that the competitors are indeed of different skills and the
result actually reflects genuine differences in abilities, rather than
the result of random chance.  Alternatively, we may view the dataset
as being sufficiently informative to reveal actual differences in
surfing ability.  Observe that the competitors' points (that is, the
final column of table~\ref{resultstable}) cannot be used for this type
of probabilistic analysis: the points awarded are arbitrary and
changing the points system would change the points scored but not
affect the competitors' behaviour.

However, points awarded may be compared with likelihood estimates, and
one would expect high points totals to be associated with high
likelihoods.  Figure~\ref{compare_likelihood_points} shows a loose
correlation ($R^2=0.71$; $p=4\times 10^{-7}$) and
Figure~\ref{compare_likelihood_points_rankings} shows how the rankings
differ when calculated by the points and likelihood systems.  Note the
anomalous position of Florence, who did not compete in five venues due
to injury: this strongly affecting his points total but not his
likelihood-based ranking.

@@

To illustrate the concept, we show a simple likelihood function in
Table~\ref{probabilitybrazilian}.  The null would be that this reified
entity has zero strength.  We fail to reject this null (differential
support = 0.035; $p=0.79$) and conclude that there is no evidence to
suggest that Brazilians' performance is better on their preferred wave
types than one would expect by chance.  We may go further using the
method of support: not only is the maximum likelihood estimate for
Brazilian wavetype very small (about 0.002), but in addition we may
reject the proposal that the strength of this entity exceeds about
0.015 (Figure~\ref{brazilian_wavetype_support}): at this value of the
strength or greater, we may gain two or more units of support by
moving to the evaluate~\citep{edwards1992}.

@@

The reified Bradley-Terry technique furnishes an intuitive statistical
test for such mechanisms, that automatically controls for competitors'
strengths.  Noting that the compatriots effectively place themselves
at a {\em disadvantage} when compared with the odd man out, we
introduce a compatriot monster whose strength $p_C$ helps the odd man
out.  Writing $p_1, p_2$ for the strengths of the compatriots and $q$
for the odd man out, we construct likelihood functions for the six
possible orderings when two compatriots and an odd man out are in the
water at the same time in Table~\ref{likelihoodcompatriot} (if we have
either three different nationalities or three identical nationalities,
we revert to the standard Plackett-Luce likelihood).  Note that all
six likelihood functions in table~\ref{likelihoodcompatriot} are
informative about $p_1,p_2,q$ in addition to being informative about
the object of inference, $p_C$.

@@

\begin{table}[h]
\begin{tabular}{l|l}
order  & ${\mathcal L}(p_1,p_2,q,p_C)$\\ \hline
$p_1\succ p_2\succ q $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2^{\vphantom{h}}}{p_2+q+p_C}$\\
$p_2\succ p_1\succ q $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{p_1^{\vphantom{h}}}{p_1+q+p_C}$\\
$p_1\succ q\succ p_2 $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{q+p_C^{\vphantom{h}}}{p_2+q+p_C}$\\
$p_2\succ q\succ p_1 $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{q+p_C^{\vphantom{h}}}{p_1+q+p_C}$\\
$q\succ p_1\succ p_2 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_1^{\vphantom{h}}}{p_1+p_2}$\\
$q\succ p_2\succ p_1 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_2^{\vphantom{h}}}{p_1+p_2}$\\
\end{tabular}
\caption{The \label{likelihoodcompatriot} six possible finishing
  orders (``$\succ$'' means ``scored more points than'') for the case
  where two compatriot surfers and an odd man out are in the water at
  the same time, with associated likelihood function for strengths
  $p_1,p_2,q, p_C$.  Here $p_1,p_2$ are the strengths of two
  compatriot surfers, $q$ the strength of the odd man out, $p_C$ the
  strength of the compatriot monster. and nationality and a third are
  in the water at the same time}
\end{table}

@@

\begin{table}[h]
\begin{tabular}{l|l|l}
wave type  & $\operatorname{Prob}(p_1\succ p_2)$ & $\operatorname{Prob}(p_2\succ p_1^{\vphantom{h}})$\\ \hline
beach & $\frac{p_1+B}{p_1+p_2^{\vphantom{h}}+B}$ & $\frac{p_2^{\vphantom{h}}}{p_1+p_2^{\vphantom{h}}+B}$\\
point & $\frac{p_1+B}{p_1+p_2^{\vphantom{h}}+B}$ & $\frac{p_2^{\vphantom{h}}}{p_1+p_2^{\vphantom{h}}+B}$\\
reef  & $\frac{p_1\vphantom{B}}{p_1+p_2}$ & $\frac{p_2^{\vphantom{h}}}{p_1^{\vphantom{h}}+p_2}$\\
\end{tabular}
\caption{Likelihood function for results between surfer 1 (Brazilian)
  and surfer 2 (non-Brazilian) with \label{probabilitybrazilian} Bradley-Terry
  strengths $p_1,p_2$ respectively, and a ``Brazilian wavetype" entity with
  strength $B$ that quantifies the enhanced Brazilian performance on reef wave types}
\end{table}
\clearpage
\newpage

\begin{table}[h]
  \begin{tabular}{l|l|l|l}\\
    \backslashbox{order}{value}          & B & F & S\\ \hline
    $\mbox{B}\succ\mbox{F}\succ\mbox{S}$ & 1 & 0 & 0\\
    $\mbox{B}\succ\mbox{S}\succ\mbox{F}$ & 1 & 0 & 1\\
    $\mbox{F}\succ\mbox{B}\succ\mbox{S}$ & 0 & 1 & 0\\
    $\mbox{F}\succ\mbox{S}\succ\mbox{B}$ & 0 & 1 & 1\\
    $\mbox{S}\succ\mbox{B}\succ\mbox{F}$ & 1 & 0 & 2\\
    $\mbox{S}\succ\mbox{F}\succ\mbox{B}$ & 0 & 1 & 2\\
  \end{tabular}
  \caption{Simplified payoff table\label{valuetable} for Buchan (B),
    Freestone (F), and Smith (S).  Buchan and Freestone, being
    Australian, have a payoff that reflects their likelihood of being
    picked for the national Olympic squad: they are interested in
    whether or not they beat their countryman.  Thus Buchan's payoff
    shows whether or not he beat Freestone, and vice versa.  Smith,
    being South African, is motivated by his ranking against his
    competitors and here we show Smith with a Borda-type payoff.
    Smith observes that {\em both} Buchan and Freestone are
    indifferent between outcomes $\left\lbrace
    \mbox{B}\succ\mbox{F}\succ\mbox{S},
    \mbox{B}\succ\mbox{S}\succ\mbox{F},
    \mbox{S}\succ\mbox{B}\succ\mbox{F}\right\rbrace$, and may be able
    to exploit this indifference by strategies increasing the
    probability of his desired outcome,
    $\mbox{S}\succ\mbox{B}\succ\mbox{F}$}
\end{table}

>>>>

\documentclass{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[comma]{natbib}
\usepackage{diagbox}
\usepackage{multirow}
\usepackage[pdftex]{hyperref}

@@

Fig.~\ref{dotchartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about $0.175$ down to Morais at about $0.01$.  The {\tt
  hyper2} software~\citep{hankin2017_nomarkup} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  This test is performed by function {\tt
  equalp.test()} in the package which uses the method of support
\citep{edwards1992}.  The test rejects the null (in favour of the
evaluate) at $p=4\times 10^{-5}$.  There is thus strong evidence to
suggest that the competitors are indeed of different skills and the
result actually reflects genuine differences in abilities, rather than
the result of random chance.  Alternatively, we may view the dataset
as being sufficiently informative to reveal actual differences in
surfing ability.  Observe that the competitors' points (that is, the
final column of Table~\ref{resultstable}) cannot be used for this type
of probabilistic analysis: the points awarded are arbitrary and
changing the points system would change the points scored but not
affect the competitors' behaviour.

However, points awarded may be compared with likelihood estimates, and
one would expect high points totals to be associated with high
likelihoods.  Fig.~\ref{compare_likelihood_points} shows a loose
correlation ($R^2=0.71$; $p=4\times 10^{-7}$) and
Fig.~\ref{compare_likelihood_points_rankings} shows how the rankings
differ when calculated by the points and likelihood systems.  Note the
anomalous position of Florence, who did not compete in five venues due
to injury: this strongly affecting his points total but not his
likelihood-based ranking.

@@

To illustrate the concept, we show a simple likelihood function in
Table~\ref{probabilitybrazilian}.  The null would be that this reified
entity has zero strength.  We fail to reject this null (differential
support = 0.035; $p=0.79$) and conclude that there is no evidence to
suggest that Brazilians' performance is better on their preferred wave
types than one would expect by chance.  We may go further using the
method of support: not only is the maximum likelihood estimate for
Brazilian wavetype very small (about 0.002), but in addition we may
reject the proposal that the strength of this entity exceeds about
0.015 (Fig.~\ref{brazilian_wavetype_support}): at this value of the
strength or greater, we may gain two or more units of support by
moving to the evaluate~\citep{edwards1992}.

@@

The reified Bradley-Terry technique furnishes an intuitive statistical
test for such mechanisms, that automatically controls for competitors'
strengths.  Noting that the compatriots effectively place themselves
at a {\em disadvantage} when compared with the odd man out, we
introduce a compatriot monster whose strength $p_C$ helps the odd man
out.  Writing $p_1, p_2$ for the strengths of the compatriots and $q$
for the odd man out, we construct likelihood functions for the six
possible orderings when two compatriots and an odd man out are in the
water at the same time in Table~\ref{likelihoodcompatriot} (if we have
either three different nationalities or three identical nationalities,
we revert to the standard Plackett-Luce likelihood).  Note that all
six likelihood functions in Table~\ref{likelihoodcompatriot} are
informative about $p_1,p_2,q$ in addition to being informative about
the object of inference, $p_C$.

@@

\begin{table}[h]
\begin{tabular}{ll}
order  & ${\mathcal L}(p_1,p_2,q,p_C)$\\ \hline
$p_1\succ p_2\succ q $ \hphantom{www}
                       & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2^{\vphantom{h}}}{p_2+q+p_C}$\\
$p_2\succ p_1\succ q $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{p_1^{\vphantom{h}}}{p_1+q+p_C}$\\
$p_1\succ q\succ p_2 $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{q+p_C^{\vphantom{h}}}{p_2+q+p_C}$\\
$p_2\succ q\succ p_1 $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{q+p_C^{\vphantom{h}}}{p_1+q+p_C}$\\
$q\succ p_1\succ p_2 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_1^{\vphantom{h}}}{p_1+p_2}$\\
$q\succ p_2\succ p_1 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_2^{\vphantom{h}}}{p_1+p_2}$\\
\end{tabular}
\caption{The \label{likelihoodcompatriot} six possible finishing
  orders (``$\succ$'' means ``scored more points than'') for the case
  where two compatriot surfers and an odd man out are in the water at
  the same time, with associated likelihood function for strengths
  $p_1,p_2,q, p_C$.  Here $p_1,p_2$ are the strengths of two
  compatriot surfers, $q$ the strength of the odd man out, $p_C$ the
  strength of the compatriot monster. and nationality and a third are
  in the water at the same time}
\end{table}

@@

\begin{table}[h]
\begin{tabular}{lll}
wave type  & $\operatorname{Prob}(p_1\succ p_2)$ & $\operatorname{Prob}(p_2\succ p_1^{\vphantom{h}})$\\ \hline
beach & $\frac{p_1+B}{p_1+p_2^{\vphantom{h}}+B}$ & $\frac{p_2^{\vphantom{h}}}{p_1+p_2^{\vphantom{h}}+B}$\\
point & $\frac{p_1+B}{p_1+p_2^{\vphantom{h}}+B}$ & $\frac{p_2^{\vphantom{h}}}{p_1+p_2^{\vphantom{h}}+B}$\\
reef  & $\frac{p_1\vphantom{B}}{p_1+p_2}$ & $\frac{p_2^{\vphantom{h}}}{p_1^{\vphantom{h}}+p_2}$\\
\end{tabular}
\caption{Likelihood function for results between surfer 1 (Brazilian)
  and surfer 2 (non-Brazilian) with \label{probabilitybrazilian} Bradley-Terry
  strengths $p_1,p_2$ respectively, and a ``Brazilian wavetype" entity with
  strength $B$ that quantifies the enhanced Brazilian performance on reef wave types}
\end{table}
\clearpage
\newpage

\begin{table}[h]
  \begin{tabular}{llll}\\
    \backslashbox{order}{value}          & B & F & S\\ \hline
    $\mbox{B}\succ\mbox{F}\succ\mbox{S}$ & 1 & 0 & 0\\
    $\mbox{B}\succ\mbox{S}\succ\mbox{F}$ & 1 & 0 & 1\\
    $\mbox{F}\succ\mbox{B}\succ\mbox{S}$ & 0 & 1 & 0\\
    $\mbox{F}\succ\mbox{S}\succ\mbox{B}$ & 0 & 1 & 1\\
    $\mbox{S}\succ\mbox{B}\succ\mbox{F}$ & 1 & 0 & 2\\
    $\mbox{S}\succ\mbox{F}\succ\mbox{B}$ & 0 & 1 & 2\\
  \end{tabular}
  \caption{Simplified payoff table\label{valuetable} for Buchan (B),
    Freestone (F), and Smith (S).  Buchan and Freestone, being
    Australian, have a payoff that reflects their likelihood of being
    picked for the national Olympic squad: they are interested in
    whether or not they beat their countryman.  Thus Buchan's payoff
    shows whether or not he beat Freestone, and vice versa.  Smith,
    being South African, is motivated by his ranking against his
    competitors and here we show Smith with a Borda-type payoff.
    Smith observes that {\em both} Buchan and Freestone are
    indifferent between outcomes $\left\lbrace
    \mbox{B}\succ\mbox{F}\succ\mbox{S},
    \mbox{B}\succ\mbox{S}\succ\mbox{F},
    \mbox{S}\succ\mbox{B}\succ\mbox{F}\right\rbrace$, and may be able
    to exploit this indifference by strategies increasing the
    probability of his desired outcome,
    $\mbox{S}\succ\mbox{B}\succ\mbox{F}$}
\end{table}
